% -*-latex-*-


%Examining the traits that govern ; Defining the cological and physiological traits of phytoplankton with metatranscriptomic approaches
\title{Defining the ecological and physiological traits of phytoplankton across marine ecosystems}

%\title{Defining ecological and physiological traits of phytoplankton with metatranscriptomics}
%\title{Illuminating the ecological and biogeochemical roles of phytoplankton through metatranscriptomics}
\author{Harriet Alexander}
\prevdegrees{B.S., Wellesley College (2010)}
% If you wish to list your previous degrees on the cover page, use the 
% previous degrees command:
%       \prevdegrees{A.A., Harvard University (1985)}
% You can use the \\ command to list multiple previous degrees
%       \prevdegrees{B.S., University of California (1978) \\
%                    S.M., Massachusetts Institute of Technology (1981)}
\department{Joint Program in Applied Ocean Science \& Engineering\\Massachusetts Institute of Technology\\ \& Woods Hole Oceanographic Institution}

% If the thesis is for two degrees simultaneously, list them both
% separated by \and like this:
% \degree{Doctor of Philosophy \and Master of Science}
\degree{Doctor of Philosophy}

% As of the 2007-08 academic year, valid degree months are September, 
% February, or June.  The default is June.
\degreemonth{February}
\degreeyear{2016}
\thesisdate{January 4, 2015}

%% By default, the thesis will be copyrighted to MIT.  If you need to copyright
%% the thesis to yourself, just specify the `vi' documentclass option.  If for
%% some reason you want to exactly specify the copyright notice text, you can
%% use the \copyrightnoticetext command.  
%\copyrightnoticetext{\copyright IBM, 1990.  Do not open till Xmas.}

\copyrightnoticetext{\copyright 2016 Harriet Alexander. All rights reserved.  
\\ The author hereby grants to MIT and WHOI permission to reproduce and 
to distribute publicly paper and electronic copies of this thesis document 
in whole or in part in any medium now known or hereafter created.}

% If there is more than one supervisor, use the \supervisor command
% once for each.
\supervisor{Sonya T. Dyhrman}{Associate Professor of Earth \& Environmental Science\\Columbia University}
\supervisor{Elizabeth B. Kujawinski}{Associate Scientist with Tenure\\Woods Hole Oceanographic Institution}

% This is the department committee chairman, not the thesis committee
% chairman.  You should replace this with your Department's Committee
% Chairman.
\chairman{Lauren Mullineaux}{Chair, Joint Committee for Biological Oceanography\\Massachusetts Institute of Technology\\Woods Hole Oceanographic Institution}

% Make the titlepage based on the above information.  If you need
% something special and can't use the standard form, you can specify
% the exact text of the titlepage yourself.  Put it in a titlepage
% environment and leave blank lines where you want vertical space.
% The spaces will be adjusted to fill the entire page.  The dotted
% lines for the signatures are made with the \signature command.
\maketitle

% The abstractpage environment sets up everything on the page except
% the text itself.  The title and other header material are put at the
% top of the page, and the supervisors are listed at the bottom.  A
% new page is begun both before and after.  Of course, an abstract may
% be more than one page itself.  If you need more control over the
% format of the page, you can use the abstract environment, which puts
% the word "Abstract" at the beginning and single spaces its text.

%% You can either \input (*not* \include) your abstract file, or you can put
%% the text of the abstract directly between the \begin{abstractpage} and
%% \end{abstractpage} commands.

% First copy: start a new page, and save the page number.
\cleardoublepage
% Uncomment the next line if you do NOT want a page number on your
% abstract and acknowledgments pages.
% \pagestyle{empty}
\setcounter{savepage}{\thepage}
\begin{abstractpage}
\input{abstract}
\end{abstractpage}

% Additional copy: start a new page, and reset the page number.  This way,
% the second copy of the abstract is not counted as separate pages.
% Uncomment the next 6 lines if you need two copies of the abstract
% page.
% \setcounter{page}{\thesavepage}
% \begin{abstractpage}
% \input{abstract}
% \end{abstractpage}

\cleardoublepage
\null\vfill
{\raggedleft\vfill\itshape\Longstack[l]{
  For my grandparents, Betty and Sherwood\bigskip\bigskip\space}\par
}
\cleardoublepage

\section*{\centerline{\LARGE\textsc{Acknowledgments}}}

{\parindent15pt
    %This thesis, like so many of my adventures to date, would not have come to fruition without the support, help, and care of the many incredible people in my life. \par

Throughout graduate school I have been supported by the MIT Presidential Fellowship, the National Defense Science and Engineering Graduate (NDSEG) Fellowship, the Ocean Life Institute Fellowship, and the WHOI Academic Programs Office. This research was supported by funds from the National Science Foundation (NSF) Environmental Genomics and NSF Biological Oceanography Programs through Grant OCE-0723667 (to S.T.D., B.D.J., and T.A.R.) and Grant OCE-0962208 (to B.D.J.), the Joint Genome Institute/Department of Energy Community Sequencing Program through Grant CSP795793 (to B.D.J., S.T.D., and T.A.R.), the Center for Microbial Oceanography: Research and Education (C-MORE, National Science Foundation Grant EF04-24599), the Gordon and Betty Moore Foundation through Grant 2637 to the National Center for Genome Resources for the MMETSP and Grant 3794 (to D.M.K.). This work was also supported through grants from the Simons Foundation (to D.M.K. and S.T.D.), and is a contribution of the Simons Collaboration on Ocean Processes and Ecology (SCOPE). \par\bigskip

The last five years has been the most interesting, challenging, and rewarding period of my life. I have been given the opportunity not only to freely pursue my academic passions, but to travel the world, interact with amazing people, and grow. Without the support and guidance of my mentors, colleagues, friends, and family, this journey would not have been possible. \par

First and foremost, I wish to thank my advisor Sonya Dyhrman, whose confidence in my scientific abilities has bolstered and motivated me throughout my thesis. Over the last five years, Sonya has nurtured my scientific curiosity challenged me to delve deeper and work harder than I thought possible. She went well beyond her prescribed duty as advisor, waking before dawn, hauling around 360L of water (more than once), and traveling to the \textit{literal} ends of the earth to make science happen. It has been a great privilege to work with such a talented scientist, whose integrity and ardent devotion to her colleagues and students I hope to emulate in my career.  \par

I am grateful to my thesis committee, Penny Chisholm, Scott Doney, Mick Follows, and Bethany Jenkins for their guidance and thoughtful input during my PhD. I especially wish to thank my co-advisor, Liz Kujawinski, who generously folded me into her lab group following Sonya's relocation and taught me the joys of FT-ICR-MS. I also would like to thank my co-authors: Tatiana Rynearson, Mak Saito, Dave Karl, Sam Wilson, and Melissa Mercier. \par 

Special thanks to the WHOI Biology Department and JCBO for the fantastic classes and welcoming environment. In particular, I would like to thank Ann Tarrant for giving me a home base at WHOI following Sonya's departure. I am also thankful to Sam Laney for his advice and guidance during the course of my PhD and his agreement to chair my defense. \par 

The vast majority of the data in this thesis were gathered at sea. I owe a big `mahalo' to the captain and crew of the R/V Kilo Moana, the R/V Knorr, and the R/V Cap'n Bert. My experiences at sea have been shaped by the wonderful people with whom I have sailed. To all the participants in the HOE-DYLAN and DeepDOM cruises, thank you for the great time and all your help sherpa-ing enormous amount of water up and down stairs.\par

I count myself lucky to have overlapped with a group of wonderful and supportive people during my time in the Dyhrman lab group. It goes without saying that Sheean Haley was the engine that keeps the lab running seamlessly. Her levity, indefatigably, and upper body strength, both on land and at sea, kept me grounded during my thesis and made all my science possible. To my fellow Dyhrman Lab graduate students and postdocs, Louie, Abby, M\'{o}nica, Solange, Colleen, and Kyle, my thanks for your camaraderie, help, and general awesomeness over the last six years. \par
I have been blessed with an amazing assortment of friends both in the Woods Hole and beyond. Thank you all for making this time fun and memorable. During this time I have been especially thankful for my cohort, who stood shoulder to shoulder with me throughout graduate school. Thank you all for the late-night pizza and P-set sessions, sympathetic shoulders, and wonderful laughs. \par

Finally, I would like to thank my mother and family for their tireless love and support, not only during my Ph.D., but throughout my life. I owe so much to my mother, who since I was small knew that I would end up working with the oceans and supported me in that dream. I wish to dedicate this thesis to the memory of my grandparents Betty and Sherwood Finley, to whom I owe my work ethic, love of nature, and perspective.\par 


}
 


